HO unification from object language to meta language

### 2024-02-21 paper/main.tex
update tex

\noindent
illustrates the \elpiIn{assmt} representing variable assignment in the ML.

Variable assignment

\vspace{30pt}

In this encoding, the fo\_tm type stands for the first order representation of
the object language terms. We model unification variables as integers
corresponding to memory addresses. The memory is represented as list of optional
terms (the subst\_fo type abbreviation). If the cell $i$ is none, then the
variable $i$ is not instantiated, otherwise it has already been assigned to some
value.

% Why the fo prefix for fo_tm && Uvar in OL have no scope
Because of our translation of the terms of the object language into the meta
language outlined in \cref{sec:intro}, the variables of the object language have 
no scope.
An illustration of this observation can be found by analyzing deeply the subterm \coqIn{P x} from
the instance forall\_dec. Here, \coqIn{P} is an higher order variables
with type \coqIn{A ~$\to$~ Prop} and
\coqIn{x} is a name bound to \coqIn{P}.
However, at the meta level, the translation of \coqIn{P x} becomes \elpiIn{app[uv
1, x]}, that is, \elpiIn{uv 1} cannot reference \elpiIn{x}.

As our objective is to handle higher-order variables, a viable solution involves
addressing this issue through the compilation of the term received as input into
a more expressive version. This can be achieved by employing a second type
structure for the terms of the OL, as outlined in the second column of
\cref{code:lang-descr}. The primary difference lies in the fact that a
unification variable (always identified by an integer) now takes a list of terms
representing its scope. The abstractions of the ML are
symbolized using the constructor \elpiIn{abs} of the type \elpiIn{assmt}. The
constructor \elpiIn{val} is utilized to contain terms of the object language,
and in this context, there is no need for an application node in the ML.

\def\eqfo{equal\_fo\xspace}
\def\eqho{equal\_ho\xspace}
\def\etabeta{$\eta\beta$\xspace}

% We have two different equal functions
The equality relation over terms under a given substitution mapping is possible 
thanks to the equal predicate. There exist
two version of this predicate: \eqfo and \eqho. Their implementation is given in
\cref{code:term-equal}.

The first tests if two terms are equal in the object language, that is, given a
substitution $\theta$, is it true that the two terms are equal? Note that no
variable instantiation is done: unification is not performed by the
\elpiIn{equal} predicate. Therefore, a unification variable $i$ equals to
another term $t$ if $t$ is a variable with same index $i$, or if $i$ is assigned
to $t'$ in $\theta$ and \elpiIn{equal ~$t'$~ ~$t$~}. Moreover, since \eqfo 
represents the equality of the OL, and since the OL accept term equality up to
\etabeta, then \eqfo also quotient over these reductions.

The \eqho predicate tests if two terms are equal in the world of the meta
languages. This equality function is not capable to understand the \etabeta
reductions over terms of the object language. The ML sees a term of the type
\elpiIn{tm} as atom on which only structural equality can be performed. It is
only possible to dereference a variable if assigned.
\marginpar{TODO: say HO is same as ML}

% \section[Compilation: fo\_tm to tm]{Compilation: \elpiIn{fo_tm} $\to$ \elpiIn{tm}}

>>>>

\noindent
illustrates the \elpiIn{assmt} representing variable assignment in the ML. \marginpar{todo: explain better}

The memory of the two languages are represented with lists of substitutions. In
particular, the substitution of the OL, called \elpiIn{fo_subst}, is made by
optional terms such that, if the substitution is \elpiIn{none}, then the
variable is not instantiated. Note that the variables in the \elpiIn{fo_subst}
have always the \elpiIn{fo_uv} constructor. On the other hand, the ML
substitution is an optional assignment and in that assignment, variables are
considered to have the \elpiIn{uv} constructor.

A key property needed in unification is being able to verify if two terms are
equal. This is kind of a structural equality verification between two terms,
where variable dereferencing is performed when the variable is assigned. A
sketch of the equality function is given for the OL language in
\cref{code:term-equal}. Though, this equality relation over terms of a language
can be powered by other reduction rules depending equational theory being
considered. In our case, the OL terms are equal under $\eta\beta$ redex. This mean
that new rules for those two redexes are added in the implementation of
\elpiIn{fo_equal}.

If \elpiIn{fo_equal} is conceived to manage equality between terms of the OL,
the same equality predicate in the ML behave slightly different. By the given
definition of the ML, the ML allows $\eta\beta$ congruence of terms, but, since
the node \elpiIn{app} and \elpiIn{abs} are constructor representing the
applications and the abstractions of the OL, these two reduction rules cannot
applied on them. We build therefore a predicate \elpiIn{equal} working for terms
in the ML which is implemented merely with the rules for the \elpiIn{fo_equal}
predicate.\todo{not true for uv} For example, if \elpiIn{fo_equal []
(abs x\ [c"f", c]) (c"f")} is true in the object language, \elpiIn{equal [] (abs
x\ [c"f", c]) (c"f")} produces a failure.

The solution we are proposing aim to overcome these unification issues by 1)
compiling the terms $t$ and $u$ of the OL into an internal version $t'$ and $u'$
in the ML; 2) unifying $t'$ and $u'$ at the meta level instantiating meta
variables; 3) decompiling the meta variable into terms of the OL; 4) assigning
the variables of the OL with the decompiled version of their corresponding meta
variables. We claim that $t$ and $u$ unify if and only if $t'$ and $u'$ unify
and that the substitution in the object language is the same as the one returned
by the ML. \todo{same or $\supseteq$ or $\subseteq$}

Mathematically, we what to prove the following property:

\begin{coqcode}
  forall (t  u : term_ol) (s : subst_ol) 
         (t' u': term_ml) (s': subst_ml), 
    equal_ol t u s <->
      comp t t' /\ comp u u' /\ 
      unif_ml t' u' s' /\ decomp s' s.
\end{coqcode}

$$Math\ formula$$

In the following section we explain how we deal with term (de)compilation and
unification variable linking.

\section[Compilation: fo\_tm to tm]{Compilation: \texttt{tm} $\to$ \texttt{tm}}
